Kindly Bent to Free Us

### 2019-11-12 kindly/infer/infer.tex
Added syntactic category for multiplicities, simplified lattice rules, fixed syntax-directed kinding rules to match syntax (applied type constructors instead of type constructors).
Fix the remaining typos/small comments from the reviews.

In \cref{sdtyping}, we used relations to split environments in two and to
transform suspended bindings into borrows inside a region.
These relations take as argument a constraint which validates
the transformations.
In the context of inference, we define new judgements which \emph{infers}
the constraints.
\begin{itemize}[leftmargin=*]
\item $\bsplit{C}{\Sv}{\bm{\Sv_1}}{\bm{\Sv_2}}$.
  Given two usage environments $\Sv_1$ and $\Sv_2$
  (inferred from two subexpressions, for instance),
  we return $\Sv$, the merged environment, and $C$, a set
  of constraints that must be respected.
  This relation is total and non-ambiguous given $\Sv_1$ and $\Sv_2$
  and uses the same rules as the one presented in \cref{sdtyping:split}.
\item $\bregion[\bm{n}]{C}{\bm{x}}{\Sv}{\bm\Sv'}$.
  Given a usage environment $\Sv'$ and a variable name $x$, we return
  $\Sv$ where the borrow binding of $x$ in $\Sv'$, if it exists, is replaced by
  a suspended binding. We also return $C$, a set of constraints that must
  be respected.
  Again, the relation is total and non-ambiguous for any given $\Sv'$ and $x$,
  and uses the same rules as the one presented in \cref{sdtyping:regions}.
\end{itemize}

@@

\subsubsection{Environments and bindings}
\label{infer:envs}
%
\begin{figure}[tp]
  \begin{mathpar}
    { \Weaken_{\bvar{x}{\sigma}}(\Sv) =
      \begin{dcases}
        \Ctrue&\text{if } x\in\Sv\\
        \Cleq{\sigma}{\kaff_\infty}&\\
      \end{dcases}
    }
    \and
    \ruleIVar
    \and
    \ruleIAbs
  \end{mathpar}
  \caption{The {\sc Abs$_I$} and {\sc Var$_I$} rules}
  \label{rule:infer:envs}
\end{figure}
%
In our syntax-directed system, we ensured
that linear variables are not discarded at the \emph{leaves}, through
the {\sc Var} rule. In the inference algorithm, we operate in the opposite
direction: we collect data from the leaves, and only enforce linearity
at \emph{binders}. This is reflected in the {\sc Var$_I$} and
{\sc Abs$_I$} rules presented in \cref{rule:infer:envs}.
Typing for variables is very similar to traditional (non-affine) ML
typing. The only difference is that we collect
the fact that $x$ was used with the scheme $\schm$ by returning
a usage environment $\Sv = \{ \bvar{x}{\schm} \}$.
%
This usage environment is in turn used at each binder to enforce proper
usage of linear variable via the $\Weaken$ property.
We demonstrate how this is achieved for lambda expressions
in the {\sc Abs$_I$} rule.
First, we typecheck the body of the lambda and obtain a usage
environment $\Sv_x$. Just like in the syntax-directed type system,
we introduce the constraint
$\Cleq{\Sv\Sdel{x}}{\kvar}$ which properly accounts for captures in
the body of the lambda expression. We also introduce the constraint
$\Weaken_{\bvar{x}{\sigma}}(\Sv)$. If $x$ was used $e$, $x$ is present
in $\Sv$ and this constraint is empty. Otherwise, if $x$ is
\emph{not} present in $\Sv$, its type should not have a linear
kind, which is enforced by the constraints $\Cleq{\sigma}{\kaff_\infty}$.
The $\Weaken$ constraint is introduced at each binding point in the program
such as let-bindings and pattern matching.

@@

    \inferrule{}
    {\generalize{C}{\E}{\tau} =
      (\Cproj{\Multi{\kvar},\Multi{\tvar}}{C},
      \forall \Multi{\kvar},\Multi{\tvar}.\qual{C}{\tau})\\
      \text{ where}\quad
      \Multi{\kvar},\Multi{\tvar} = (\fv{\tau}\cup\fv{C})\setminus\fv{\E}
    }
  \end{mathpar}
  \caption{The {\sc Let$_I$} rule}
  \label{rule:infer:let}
\end{figure}
%
So far, constraints have only been composed of a list of kind inequalities.
The internal language of constraints used for inference is in fact
richer.
To illustrate this, we show the inference rule for let-bindings
in \cref{rule:infer:let}.
This rule combines several ingredients previously seen:
since let expressions are binders, we use $\Weaken$ on the bound
identifier $x$. Since let-expressions contain
two subexpressions, we use the environment splitting relation,
$\bsplit{C_s}{\Sv}{\Sv_1}{(\Sv_2 \Sdel{x})}$. We remove the $x$ from
the right environment, since it is not available outside of the expression
$e_2$, and should not be part of the returned usage environment.

As per tradition in ML languages, generalization is performed
on let-bindings.
Following HM(X), we write $(C_\schm, \schm) = \generalize{C}{\E}{\tau}$
for the pair of a constraint and a scheme resulting from
generalization. The definition is provided in \cref{rule:infer:let}.
The type scheme $\schm$ is created by quantifying over all the appropriate
free variables and the current constraints.
The generated constraint $C_\schm$ uses a new projection operator,
$\Cproj{\tvar}{D}$, which
allow the creation of local variables inside the constraints.
This allows us to encapsulate all the quantified variables in the global constraints.
It also reflects the fact that there
should exist at least one solution for $C$ for the scheme to be valid.
\citet{DBLP:journals/tapos/OderskySW99} give a detailed account
on the projection operators in HM inference.

@@

We define the set of solved formed
$\mathcal S$ as the set of constraints $\CL$ quantified by $\equivC$.
We will show later that such constraints are in fact only composed of
kind inequalities, and thus correspond to the syntactic constraints
used in type and kind schemes.
%
We now define our normalization procedure $\normalize{C_0}{\unif_0}$, where
$C_0\in \CL$ is a set of constraints and $\unif_0$ is a substitution.
It returns a constraint $C \in \mathcal S$ in
solved form and a unifier $\unif$.
The main idea of the algorithm is to first remove all the type equalities
by using regular Herbrand unification. After that, we only have
a set of inequalities among kinds, which we can consider as a relation.
We can then saturate the relation,
unify all kinds that are in the same equivalence classes to obtain
a most general unifier on kind variables,
remove all existentially quantified variables and
then minimize back the relation and apply various
simplification rules to make the resulting type easier to understand to users.

More precisely, we apply the following steps:
\begin{enumerate}
\item Solve all type equality constraints through Herband unification and
  gather all existential quantifications at the front of the constraint.
  We obtain a constraint $C^k = \exists \kvar_i,\ \Cleq{k_j}{k'_j}_j$ and
  a substitution $\unif_\tau$.

@@

\item For each strongly connected component in $\mathcal G$, unify all its vertices and replace it by a representative.
  We write $\unif_k$ for the substitution that replaces a kind variable by
  its representative.
  The representative of a strongly connected component $g$ can be determined as follows:
  \begin{itemize}
  \item If $g$ does not contain any constant, then the representative
    is a fresh kind variable.
  \item If $g$ contains exactly one constant, it is the representative.
  \item Otherwise, the initial constraint $C_0$ is not satisfiable.
  \end{itemize}
  Note that this step will also detect all unsatisfiable constraints.
\item Take the transitive closure of $\mathcal R$.
\item Remove all the vertices corresponding to the kind variables $\kvar_i$
  that are existentially quantified in $C^k$.
\item Take the transitive reduction of $\mathcal R$.
\item Remove the extremums of $\mathcal L$ and the edges of $(\leq)$
  from $\mathcal R$.
\item Return $C = \left\{ k \leq k' \mid k \operatorname{\mathcal R}k' \right\}$
  and $\unif =  \unif_\tau \meet \unif_k$.
\end{enumerate}

@@

\begin{definition}[Instance relation]
  Given a constraints $C$ and two schemes
  $\schm = \forall \Multi\tvar. \qual{D}{\tau}$ and
  $\schm' = \forall \Multi\tvar'. \qual{D'}{\tau'} $.
  Then $\entail{C}{\schm \preceq \schm'}$
  iff $\entail{C}{\subst{\tvar}{\tau''} D}$
  and $\entail{C\Cand D'}{\Cleq{\subst{\tvar}{\tau''}\tau}{\tau'}}$
\end{definition}

>>>>

In \cref{sdtyping}, we used relations to split environments in two and to
transform suspended bindings into borrows inside a region.
These relations take as argument a constraint which validates
the transformations.
In the context of inference, we define new judgements which \emph{infer}
the constraints.
\begin{itemize}[leftmargin=*]
\item $\bsplit{C}{\Sv}{\bm{\Sv_1}}{\bm{\Sv_2}}$.
  Given two usage environments $\Sv_1$ and $\Sv_2$
  (inferred from two subexpressions, for instance),
  we return $\Sv$, the merged environment, and $C$, a set
  of constraints that must be respected.
  This relation is total and non-ambiguous given $\Sv_1$ and $\Sv_2$
  and uses the same rules as the one presented in \cref{sdtyping:split}.
\item $\bregion[\bm{n}]{C}{\bm{x}}{\Sv}{\bm\Sv'}$.
  Given a usage environment $\Sv'$ and a variable name $x$, we return
  $\Sv$ where the borrow binding of $x$ in $\Sv'$, if it exists, is replaced by
  a suspended binding. We also return $C$, a set of constraints that must
  be respected.
  Again, the relation is total and non-ambiguous for any given $\Sv'$ and $x$,
  and uses the rules presented in \cref{sdtyping:regions}.
\end{itemize}

@@

\subsubsection{Environments and bindings}
\label{infer:envs}
%
\begin{figure}[tp]
  \begin{mathpar}
    { \Weaken_{\bvar{x}{\sigma}}(\Sv) =
      \begin{dcases}
        \Ctrue&\text{if } x\in\Sv\\
        \Cleq{\sigma}{\kaff_\infty}&\\
      \end{dcases}
    }
    \and
    \ruleIVar
    \and
    \ruleIAbs
  \end{mathpar}
  \caption{The {\sc Abs$_I$} and {\sc Var$_I$} rules}
  \label{rule:infer:envs}
\end{figure}
%
In our syntax-directed system, we ensured
that linear variables are not discarded at the \emph{leaves}, through
the {\sc Var} rule. In the inference algorithm, we operate in the opposite
direction: we collect data from the leaves, and only enforce linearity
at \emph{binders}. This is reflected in the {\sc Var$_I$} and
{\sc Abs$_I$} rules presented in \cref{rule:infer:envs}.
Typing for variables is very similar to traditional (non-affine) ML
typing. The only difference is that we collect
the fact that $x$ was used with the scheme $\schm$ by returning
a usage environment $\Sv = \{ \bvar{x}{\schm} \}$.
%
This usage environment is in turn used at each binder to enforce proper
usage of linear variable via the $\Weaken$ property.
We demonstrate how this is achieved for lambda expressions
in the {\sc Abs$_I$} rule.
First, we typecheck the body of the lambda and obtain a usage
environment $\Sv_x$. Just like in the syntax-directed type system,
we introduce the constraint
$\Cleq{\Sv\Sdel{x}}{\kvar}$ which properly accounts for captures in
the body of the lambda expression. We also introduce the constraint
$\Weaken_{\bvar{x}{\sigma}}(\Sv)$, which is false only if weakening
is necessary but impossible due to linearity.
The $\Weaken$ constraint is introduced at each binding point in the program
such as let-bindings and pattern matching.

@@

    \inferrule{}
    {\generalize{C}{\E}{\tau} =
      (\Cproj{(\Multi{\kvar},\Multi{\tvar})}{C},
      \forall (\Multi{\kvar},\Multi{\tvar}).\qual{C}{\tau})\\
      \text{ where}\quad
      \Multi{\kvar},\Multi{\tvar} = (\fv{\tau}\cup\fv{C})\setminus\fv{\E}
    }
  \end{mathpar}
  \caption{The {\sc Let$_I$} rule}
  \label{rule:infer:let}
\end{figure}
%
So far, constraints have only been composed of a list of kind inequalities.
The internal language of constraints used for inference is in fact
richer.
To illustrate this, we show the inference rule for let-bindings
in \cref{rule:infer:let}.
This rule combines several ingredients previously seen:
since let expressions are binders, we use $\Weaken$ on the bound
identifier $x$. Since let-expressions contain
two subexpressions, we use the environment splitting relation,
$\bsplit{C_s}{\Sv}{\Sv_1}{(\Sv_2 \Sdel{x})}$. We remove the $x$ from
the right environment, since it is not available outside of the expression
$e_2$, and should not be part of the returned usage environment.

As per tradition in ML languages, generalization is performed
on let-bindings.
Following HM(X), we write $(C_\schm, \schm) = \generalize{C}{\E}{\tau}$
for the pair of a constraint and a scheme resulting from
generalization. The definition is provided in \cref{rule:infer:let}.
The type scheme $\schm$ is created by quantifying over all the appropriate
free variables and the current constraints.
The generated constraint $C_\schm$ uses a new projection operator,
$\Cproj{x}{D}$ where $x$ can be either a type or a kind variable, which
allow the creation of local variables inside the constraints.
This allows us to encapsulate all the quantified variables in the global constraints.
It also reflects the fact that there
should exist at least one solution for $C$ for the scheme to be valid.
\citet{DBLP:journals/tapos/OderskySW99} give a detailed account
on the projection operators in HM inference.

@@

We define the set of solved formed
$\mathcal S$ as the quotient set of $\CL$ by $\equivC$.
We will show later that such constraints are in fact only composed of
kind inequalities, and thus correspond to the syntactic constraints
used in type and kind schemes.
%
We now define our normalization procedure $\normalize{C_0}{\unif_0}$, where
$C_0\in \CL$ is a set of constraints and $\unif_0$ is a substitution.
It returns a constraint $C \in \mathcal S$ in
solved form and a unifier $\unif$.
The main idea of the algorithm is to first remove all the type equalities
by using regular Herbrand unification. After that, we only have
a set of inequalities among kinds, which we can consider as a relation.
We can then saturate the relation,
unify all kinds that are in the same equivalence classes to obtain
a most general unifier on kind variables,
remove all existentially quantified variables and
then minimize back the relation and apply various
simplification rules to make the resulting type easier to understand to users.

More precisely, we apply the following steps:
\begin{enumerate}
\item Solve all type equality constraints through Herband unification and
  gather all existential quantifications at the front of the constraint.
  We obtain a constraint $C^k = \exists \Multi\kvar,\ \Cleq{k_j}{k'_j}_j$ and
  a substitution $\unif_\tau$.

@@

\item For each strongly connected component in $\mathcal G$, unify all its vertices and replace it by a representative.
  We write $\unif_k$ for the substitution that replaces a kind variable by
  its representative.
  The representative of a strongly connected component $g$ can be determined as follows:
  \begin{itemize}
  \item If $g$ does not contain any constant, then the representative
    is a fresh kind variable.
  \item If $g$ contains exactly one constant, it is the representative.
  \item Otherwise, the initial constraint $C_0$ is not satisfiable.
  \end{itemize}
  Note that this step will also detect all unsatisfiable constraints.
\item Take the transitive closure of $\mathcal R$.
\item Remove all the vertices corresponding to the kind variables $\Multi\kvar$
  that are existentially quantified in $C^k$.
\item Take the transitive reduction of $\mathcal R$.
\item Remove the extremums of $\mathcal L$ and the edges of $(\leq)$
  from $\mathcal R$.
\item Return $C = \left\{ k \leq k' \mid k \operatorname{\mathcal R}k' \right\}$
  and $\unif =  \unif_\tau \meet \unif_k$.
\end{enumerate}

@@

\begin{definition}[Instance relation]
  Given a constraint $C$ and two schemes
  $\schm = \forall \Multi\tvar. \qual{D}{\tau}$ and
  $\schm' = \forall \Multi\tvar'. \qual{D'}{\tau'} $.
  Then $\entail{C}{\schm \preceq \schm'}$
  iff $\entail{C}{\subst{\tvar}{\tau''} D}$
  and $\entail{C\Cand D'}{\Cleq{\subst{\tvar}{\tau''}\tau}{\tau'}}$
\end{definition}

### 2019-11-12 kindly/lang.tex
Fix the remaining typos/small comments from the reviews.

\lang is a minimal ML-like language with
let-polymorphism, abstract types, and pairs. Its type system manages
linearity and borrows using kinds and (kind-) constrained types.
For ease of presentation, we keep the language as simple as possible.
\begin{itemize}
\item Pattern matching is demonstrated on pairs rather than algebraic
  datatypes.
\item There are separate operators for borrowing and reborrowing (taking
  the borrow of a borrow).
\item Regions are assumed to be explicit in the code. (But we present
  an algorithm to infer them in \cref{regionannot}.)
\item Regions are identified using de Bruijn-indexing instead of region
  variables.
\end{itemize}

@@

A type $\tau$ is either a type variable, a function type indexed by a
kind,
a type application $\tapp{\tcon}{\Multi{\tau}}$ of an abstract type constructor
$\T t$, or a borrowed type  $\borrowty{k}{\tau}$.

@@

% The syntax of \lang is presented in \cref{grammar} and follows
% traditional ML languages with let bindings.
% Borrows $\borrow{e}$ are annotated with a specification indicating
% if a borrow is immutable ($\IBORROW$) or mutable ($\MBORROW$).
% We also distinguish reborrows, which are written as $\reborrow{e}$.
% Match are indexed with a specification which can be either $\&^b$
% (which $b$ a borrow specification) or $id$, indicating if the match
% should operate on borrows or not.
% Regions $\region{\Sone x\BORROW}{e}$ are annotated with their nesting ($n$)
% and the variable whose borrow they enclose ($x$).
% %
% $\tapp{\tcon}{\Multi{\tau}}$ denotes a type application where
% $\T t$ is an abstract type constructor.
% Arrow and borrow types are annotated with a kind, written as $\tau \tarr{k} \tau$
% and $\borrowty{k}{\tau}$ respectively.
% %
% Kinds can be either a kind variable $\kvar$, or constants
% (linear $\klin$, affine $\kaff$ or unrestricted $\kun$) indexed
% by a nesting level represented by a member of $N \cup \{\infty\}$.
% Type schemes and kind schemes are guarded by a set of constraints.
% %
% Finally, constraints are composed of a list of kind inequalities.

@@

A storable $W$ is a particular object that can be stored. There are five
kinds of storable. A poly closure represents a polymorphic
function. It consists of an environment and the components of an
elaborated abstraction. Similarly, a closure represents a monomorphic
function and it elides kind variables and the constraint. A pair
consists of a kind and two results. A resource is just an indirection
to another result and the $\blob $ fills a released location.

>>>>

\lang is a minimal ML-like language with
let-polymorphism, abstract types, and pairs. Its type system manages
linearity and borrows using kinds and (kind-) constrained types.
For ease of presentation, we keep the language as simple as possible.
\begin{itemize}
\item Pattern matching is demonstrated on pairs rather than algebraic
  datatypes.
\item There are separate operators for borrowing and reborrowing (taking
  the borrow of a borrow).
\item Regions are assumed to be explicit in the code. (But we present
  an algorithm to infer them in \cref{regionannot}.)
\item Regions are identified using nesting levels instead of region
  variables.
\end{itemize}

@@

A type $\tau$ is either a type variable, a function type indexed by a
kind,
a type application $\tapp{t}{\Multi{\tau}}$ of an abstract type constructor
$\T t$, or a borrowed type  $\borrowty{k}{\tau}$.

@@

% The syntax of \lang is presented in \cref{grammar} and follows
% traditional ML languages with let bindings.
% Borrows $\borrow{e}$ are annotated with a specification indicating
% if a borrow is immutable ($\IBORROW$) or mutable ($\MBORROW$).
% We also distinguish reborrows, which are written as $\reborrow{e}$.
% Match are indexed with a specification which can be either $\&^b$
% (which $b$ a borrow specification) or $id$, indicating if the match
% should operate on borrows or not.
% Regions $\region{\Sone x\BORROW}{e}$ are annotated with their nesting ($n$)
% and the variable whose borrow they enclose ($x$).
% %
% $\tapp{t}{\Multi{\tau}}$ denotes a type application where
% $\T t$ is an abstract type constructor.
% Arrow and borrow types are annotated with a kind, written as $\tau \tarr{k} \tau$
% and $\borrowty{k}{\tau}$ respectively.
% %
% Kinds can be either a kind variable $\kvar$, or constants
% (linear $\klin$, affine $\kaff$ or unrestricted $\kun$) indexed
% by a nesting level represented by a member of $N \cup \{\infty\}$.
% Type schemes and kind schemes are guarded by a set of constraints.
% %
% Finally, constraints are composed of a list of kind inequalities.

@@

A storable $w$ is a particular object that can be stored. There are five
kinds of storable. A poly closure represents a polymorphic
function. It consists of an environment and the components of an
elaborated abstraction. Similarly, a closure represents a monomorphic
function and it elides kind variables and the constraint. A pair
consists of a kind and two results. A resource is just an indirection
to another result and the $\blob $ fills a released location.

### 2019-11-12 kindly/motivation.tex
Fix the remaining typos/small comments from the reviews.

Session typing \cite{Honda1993,DBLP:conf/esop/HondaVK98} is a type
discipline for checking protocols statically. A session type is
ascribed to a communication channel and describes
a sequence of interactions. For instance, the type
\lstinline{!int.!int.?int.end} specifies a protocol
where the program must send two integers, receive an integer, and then
close the channel.
%
In this context, channels must be used linearly, because every use of
a channel ``consumes'' one interaction and  changes the type of the
channel. That is, after sending two integers on the channel,
the remaining channel has type \lstinline{?int.end}.

>>>>

Session typing \cite{Honda1993,DBLP:conf/esop/HondaVK98} is a type
discipline for checking protocols statically. A session type is
ascribed to a communication channel and describes
a sequence of interactions. For instance, the type
\lstinline{int!int!int?end} specifies a protocol
where the program must send two integers, receive an integer, and then
close the channel.
%
In this context, channels must be used linearly, because every use of
a channel ``consumes'' one interaction and  changes the type of the
channel. That is, after sending two integers on the channel,
the remaining channel has type \lstinline{int?end}.

### 2019-11-12 kindly/soa.tex
Fix the remaining typos/small comments from the reviews.

Quill~\citep{DBLP:conf/icfp/Morris16} is an Haskell-like language with affine
types through the use of qualified types.
Quill does not expose a kind language, but
uses typeclass-like mechanism and annotation on arrows.
It supports both a most general type for function composition and
principal type inference.
We were able to express the type signatures of all Quill examples
in \lang by leveraging kind inequalities.
Unlike Quill,
we provide a linearity-aware semantics and prove its soundness.
Quill does not support borrows.

@@

Alms~\citep{DBLP:conf/popl/TovP11} is an ML-like language with rich, kind-based
affine types and ML modules, similar to \lang.
Their system often relies on existential types to track the identity
of objects. For instance
\lstinline/Array.create : int -> 'a -> \E 'b. ('a, 'b) array/ where
\lstinline/'b/ uniquely identify of the array.
Due to the reliance on existentials, Alms does not support complete type inference.
Furthermore, Alms does not support borrow and often rely
on manual passing of capabilities.
In our experience, \affe's limited support for existential types through
regions is sufficient to express Alms' examples and leads to
a more convenient programming style for imperative code.
%
Kind-wise, Alms relies
on unions, intersections and dependent kinds while
\lang uses qualified types.
We believe most of Alms' kind signatures can be expressed equivalently in
our system: for instance the pair type constructor
has kind $\Pi\alpha\Pi\beta. \langle\alpha\rangle \sqcup \langle\beta\rangle$
(where $\alpha$ and $\beta$ are types and $\Pi$ is the dependent function)
in Alms and $\kvar\to\kvar\to\kvar$ in \lang thanks
to subkinding.
%
Finally, Alms provides excellent support for abstraction through
modules by allowing to keep some type unrestricted inside a module, but
exposing it as affine. We support a similar programming style, which
we showcase in \cref{motivation}.

The goal of Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
(LH) is to retrofit linear types to Haskell.
LH introduces a new notion
of ``linear arrows'', written $\multimap$ similar to linear logic,
which are functions that \emph{use} their argument exactly once.
% By contrast, in \lang linearity is only decided
% by the kinds, and $\tarr{\klin}$ simply denotes \emph{single-use functions}.
This design is easy to retrofit on top of an existing compiler
such as GHC, but there are quite controversial discussions about it\footnote{
  See the in-depth discussion attached to the GHC proposal starting here: \url{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}}.
Most relevant to \lang:
\begin{itemize}[leftmargin=*]
\item LH does not admit subtyping for arrows and requires
  $\eta$-expansion to pass unrestricted functions in linear
  contexts. This approach is acceptable in a non-strict language such as
  Haskell but changes the semantics in a strict setting.
\item
  The GHC implementation of LH performs type inference, but
  it is neither formalized nor complete.
\item
  LH promotes a continuation-passing style with functions such as
  \lstinline/withFile : path -> (file ->. file) ->. unit/
  to ensure linear use of resources. This style leads to problems with
  placing the annotation on, e.g., the IO monad.
  \lang follows System F\degree, Quill, and Alms, all of which support
  resource handling in direct style, where types themselves are
  described as affine or linear. (Or course, continuation-passing
  style is also supported.)
  We expect that the direct approach eases modular reasoning about linearity.
  In particular, using abstraction through modules,
  programmers only need to consider the module
  implementation to ensure that linear
  resources are properly handled.
% \item
%   Linear Haskell introduces a notion of linear monad to express
%   imperative code conveniently. Again, this solution is suitable in Haskell,
%   but less appropriate in other contexts. Our borrow system allow
%   to provide a more free-form programming style for imperative code.
% \item
%   Due to the previous remark, ensuring that a given object is never
%   aliased is fairly difficult in Linear Haskell, as one would need to ensure
%   that no non-linear function can manipulate it. On the other
\end{itemize}

@@

\subsection{Type-system features}
%
\lang relies on constrained types
to introduce the kind inequalities required for linear types. In
particular, we use HM(X)~\citep{DBLP:journals/tapos/OderskySW99} which
allows to use constrained types in an ML-like language with complete
type inference.
HM(X) has been shown to be compatible with subtyping,
bounded quantification and existentials~\citep{DBLP:conf/icfp/Simonet03},
GADTs~\citep{DBLP:journals/toplas/SimonetP07},
and there exists a syntactic soundness proof~\citep{DBLP:journals/entcs/SkalkaP02}.
These results make us confident that the system developed in \lang
could be applied to larger and more complex languages such as OCaml.

@@

The subtyping discipline used in \lang is very similar
to structural subtyping, where the only subtyping (or here, subkinding)
is at the leaves.
Such discipline is known to be friendly to inference and has been used in many
contexts, including OCaml itself, and has been combined
with constraints~\citep{DBLP:journals/tapos/OderskySW99,DBLP:conf/sas/TrifonovS96}.
In particular, Flowcaml~\citep{DBLP:conf/popl/PottierS02}
extends OCaml with security levels forming a lattice and supports type inference.
In \lang, we repurpose these ideas to check for linearity and affinity.
\citet{DBLP:conf/aplas/Simonet03} also presents various algorithmic
notions to solve lattice-based constraints, which we partially use
in our constraint solving algorithm. Our main contribution
is the design of a kind language specialized for checking linearity
and sufficiently simple to make
all simplification rules complete, which allows to keep type signatures simple.

>>>>

Quill~\citep{DBLP:conf/icfp/Morris16} is a Haskell-like language with affine
types through the use of qualified types.
Quill does not expose a kind language, but
uses typeclass-like mechanism and annotation on arrows.
It supports both a most general type for function composition and
principal type inference.
We were able to express the type signatures of all Quill examples
in \lang by leveraging kind inequalities.
Unlike Quill,
we provide a linearity-aware semantics and prove its soundness.
Quill does not support borrows.

@@

Alms~\citep{DBLP:conf/popl/TovP11} is an ML-like language with rich, kind-based
affine types and ML modules, similar to \lang.
Their system often relies on existential types to track the identity
of objects. For instance
\lstinline/Array.create : int -> 'a -> \E 'b. ('a, 'b) array/ where
\lstinline/'b/ uniquely identify the array.
Due to the reliance on existentials, Alms does not support complete type inference.
Furthermore, Alms does not support borrow and often rely
on manual passing of capabilities.
In our experience, \affe's limited support for existential types through
regions is sufficient to express Alms' examples and leads to
a more convenient programming style for imperative code.
%
Kind-wise, Alms relies
on unions, intersections and dependent kinds while
\lang uses qualified types.
We believe most of Alms' kind signatures can be expressed equivalently in
our system: for instance the pair type constructor
has kind $\Pi\alpha\Pi\beta. \langle\alpha\rangle \sqcup \langle\beta\rangle$
(where $\alpha$ and $\beta$ are types and $\Pi$ is the dependent function)
in Alms and $\kvar\to\kvar\to\kvar$ in \lang thanks
to subkinding.
%
Finally, Alms provides excellent support for abstraction through
modules by allowing to keep some type unrestricted inside a module, but
exposing it as affine. We support a similar programming style, which
we showcase in \cref{motivation}.

The goal of Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
(LH) is to retrofit linear types to Haskell.
LH introduces a new notion
of ``linear arrows'', written $\multimap$ similar to linear logic,
which are functions that \emph{use} their argument exactly once.
% By contrast, in \lang linearity is only decided
% by the kinds, and $\tarr{\klin}$ simply denotes \emph{single-use functions}.
This design is easy to retrofit on top of an existing compiler
such as GHC, but there are quite controversial discussions about it\footnote{
  See the in-depth discussion attached to the GHC proposal starting here: \url{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}}.
Most relevant to \lang:
\begin{itemize}[leftmargin=*]
\item LH does not admit subtyping for arrows and requires
  $\eta$-expansion to pass unrestricted functions in linear
  contexts. This approach is acceptable in a non-strict language such as
  Haskell but changes the semantics in a strict setting.
\item
  The GHC implementation of LH performs type inference, but
  it is neither formalized nor complete.
\item
  LH promotes a continuation-passing style with functions such as
  \lstinline/withFile : path -> (file ->. file) ->. unit/
  to ensure linear use of resources. This style leads to problems with
  placing the annotation on, e.g., the IO monad.
  \lang follows System F\degree, Quill, and Alms, all of which support
  resource handling in direct style, where types themselves are
  described as affine or linear. (Of course, continuation-passing
  style is also supported.)
  We expect that the direct approach eases modular reasoning about linearity.
  In particular, using abstraction through modules,
  programmers only need to consider the module
  implementation to ensure that linear
  resources are properly handled.
% \item
%   Linear Haskell introduces a notion of linear monad to express
%   imperative code conveniently. Again, this solution is suitable in Haskell,
%   but less appropriate in other contexts. Our borrow system allow
%   to provide a more free-form programming style for imperative code.
% \item
%   Due to the previous remark, ensuring that a given object is never
%   aliased is fairly difficult in Linear Haskell, as one would need to ensure
%   that no non-linear function can manipulate it. On the other
\end{itemize}

@@

\subsection{Type-system features}
%
\lang relies on constrained types
to introduce the kind inequalities required for linear types. In
particular, we use HM(X)~\citep{DBLP:journals/tapos/OderskySW99} which
allows us to use constrained types in an ML-like language with complete
type inference.
HM(X) has been shown to be compatible with subtyping,
bounded quantification and existentials~\citep{DBLP:conf/icfp/Simonet03},
GADTs~\citep{DBLP:journals/toplas/SimonetP07},
and there exists a syntactic soundness proof~\citep{DBLP:journals/entcs/SkalkaP02}.
These results make us confident that the system developed in \lang
could be applied to larger and more complex languages such as OCaml.

@@

The subtyping discipline used in \lang is very similar
to structural subtyping, where the only subtyping (or here, subkinding)
is at the leaves.
Such discipline is known to be friendly to inference and has been used in many
contexts, including OCaml, and has been combined
with constraints~\citep{DBLP:journals/tapos/OderskySW99,DBLP:conf/sas/TrifonovS96}.
In particular, Flowcaml~\citep{DBLP:conf/popl/PottierS02}
extends OCaml with security levels forming a lattice and supports type inference.
In \lang, we repurpose these ideas to check for linearity and affinity.
\citet{DBLP:conf/aplas/Simonet03} also presents various algorithmic
notions to solve lattice-based constraints, which we partially use
in our constraint solving algorithm. Our main contribution
is the design of a kind language specialized for checking linearity
and sufficiently simple to make
all simplification rules complete, which allows us to keep type signatures simple.
